Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-03-27 endofquarterpaper.qmd
editing discussion

artifact_count_median <- median(artifact_count_df$artifact_count)
n_material <- c(unique(signbase_full_clean$material))

@@

The geographic structure of the groups changes from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups. The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together. In the Evolved Aurignacian, we see a wider distribution and more distance between sites.

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994]. It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose. However, without examining environmental data, this remains only a hypothesis at this time.

@@

These results are supported by the existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. This is supported by other research finding similar evidence of artistic pro

<!-- Add evidence about scholars seeing expansion in signs/artwork following population expansions in other regions/time periods-->

Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

Our work adds to a growing awareness of diversity within the Upper Paleolithic. Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006]. Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @belfer-cohen2003; @bordes1970; @bordes1963]. This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries, as well as patterns of exchange and movement between groups.

This work also adds to our knowledge on connectivity within the Upper Paleolithic. Our network analysis and perMANOVA test results show evolving patterns of connectivity between groups and sites, suggesting dynamic patterns of social exchange and connection.

This study also adds to the growing body of methods for the identification of cultural groups, as well as change in social dynamics over time. We have shown that seriation and network analysis can be successfully used to identify cultural groups from geometric signs on mobile objects, and modularity and perMANOVA tests can be used to evaluate the strength of those groupings.

One limitation of the study is the applicability of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups. It is worth considering whether occurrence of those simpler signs at different sites is strong evidence of cultural connection, as the simplicity of the signs means an increase in the possibility that they could have been independently invented, without any sort of cultural connection of exchange. Ultimately, we chose to evaluate them as evidence of cultural connection, however, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.

>>>>

artifact_count_median <- median(artifact_count_df$artifact_count)

n_material <- c(unique(signbase_full_clean$material))

@@

The geographic boundaries of the groups change from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups. The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together. In the Evolved Aurignacian, we see a wider distribution and more distance between sites.

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons. Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

. It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose. However, without examining environmental data, this remains only a hypothesis at this time.

@@

These results are supported by the existing data on the environmental and biological change established during these phases. @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP. This rapid expansion could be what stimulated the emergence of signs as a marker of group membership. Other studies have also found a similar associations between quantitity and diversity of art, and population density increases in other regions and time periods [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992]. Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations. This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance. Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions. @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian. Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.

Our work adds to a growing awareness of diversity and interconnection within the Upper Paleolithic. Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006]. Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Belfer-Cohen_Goring-Morris_2002; @Maher_Conkey_2019; @Vanhaeren_d; @belfer-cohen2003; @bordes1970; @Shemer_Barzilai_Marder_2024; @Schwendler_2012; @bordes1963]. This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries.

This study adds to our knowledge of social networks within the Aurignacian period, as our network analysis and perMANOVA test results suggest dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of inter-group complexity at around 30 k BP, however, our analyses show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of inter-connectivity than previously predicted. We hope that dividing the data up into smaller time periods serves to rectify some of the issues in low temporal resolution that tend to affect results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023]. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].

One limitation of the study is the inclusion of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups. It is worth considering whether occurrence of those simpler signs at different sites is strong evidence of cultural connection, as the simplicity of the signs means an increase in the possibility that they could have been independently invented, without any sort of cultural connection of exchange. Ultimately, we chose to include them as evidence of cultural connection, as a greater number of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023]. However, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.
